(Deep) Induction for GADTs

### 2021-04-07 induction-paper/GADT-induction.tex
change foralls to lambdas in section 4.3
Add some notes
define KBool, K1, etc. to make lambda term examples and Section 6 complete

GADTs are expressive enough to represent, for example, the types and terms of the simply typed lambda calculus. 
We will use a variant of the STLC including a base type of Bool, and type constructors for arrow types and list types:

@@

The term calculus for this type system includes variables, abstraction, application, 
and also includes an introduction rule that takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.

\begin{equation}\label{eq:eq_lterm}
\begin{array}{l}
\mathsf{data\, LTerm\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Var\,\,\,\,\,\,:\, String \to LType\,a \to LTerm\,a} \\
\mathsf{\;\;\;\;\;\;\;\;Abs\,\,\,\,\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \to c) \to String 
      \to LType\,b \to LTerm\,c \to LTerm\,a}\\
  \mathsf{\;\;\;\;\;\;\;\;App\,\,\,\,\, :\, \exists (b : Set).\,LTerm (b \to a) \to LTerm\,b \to LTerm\,a}\\
  \mathsf{\;\;\;\;\;\;\;\;ListC\,\, :\, \exists (b : Set).\,Equal\,a\,(\List\,b) \to List\,(LTerm\,b) \to LTerm\,a}
\end{array}
\end{equation}

The type parameter for $\mathsf{LTerm}$ tracks the types of STLC terms. So the type $\mathsf{LTerm\,A}$ 
contains lambda terms of type $\mathsf{A}$. The $\mathsf{LType}$ GADT is used to encode the 
grammar of permissible types for this calculus.

Variables are represented as strings and are tagged with their type. 
The $\mathsf{Abs}$ constructor takes a name (string) and a type for the bound variable,
and a lambda term of type $\mathsf{c}$ for the body of the abstraction.
The $\mathsf{App}$ constructor applies a lambda term of type $\mathsf{b \to a}$ to a lambda term of type $\mathsf{b}$, producing
a lambda term of type $\mathsf{a}$. 
The $\mathsf{ListC}$ constructor takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.

@@

With these liftings, we can define the deep induction princple for $\mathsf{LTerm}$.
% Var case 
Let CVar be the type 
\begin{multline*}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
\mathsf{\to 
  \forall (a : Set) (Q_a : a \to Set) (s : String) (T_a : LType\, a) 
  \to LType^{\wedge} \, a\, Q_a\, T_a
  \to P \, a\, Q_a\, (Var \, s\, T_a)
  }
\end{multline*}
associated to the $\mathsf{Var}$ constructor. 
% Abs case 
Let CAbs be the type 
\begin{align*}
  &\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
  &\mathsf{\to 
  \forall (a \,b\, c: Set) (Q_a : a \to Set)  (Q_b : b \to Set) (Q_c : c \to Set)
  (e : Equal\, a\, (b \to c)) (s : String) } \\ 
  &\mathsf{ \to (T_b : LType\, b) \to (t_c : LTerm\, c)
  \to Equal^{\wedge}\,a\,(b \to c)\, Q_a \, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c) \, e
  } \\
  &\mathsf{
  \to LType^{\wedge}\, b\, Q_b\, T_b
  \to P\, c\, Q_c\, t_c\, 
  \to P \, a\, Q_a\, (Abs \,b \,c \, e \,s \,T_b \, t_c)
  }
\end{align*}
associated to the $\mathsf{Abs}$ constructor. 
% App case 
Let CApp be the type 
\begin{align*}
  &\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
  &\mathsf{\to 
  \forall (a \,b : Set) (Q_a : a \to Set)  (Q_b : b \to Set) 
   (t_{ba} : LTerm\, (b \to a)) (t_b : LTerm\, b)} \\
  &\mathsf{
  \to P\, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a) \, t_{ba} \, 
  \to P\, b\, Q_b\, t_b\, 
  \to P \, a\, Q_a\, (App \,b \,t_{ba} \, t_b) }
\end{align*}
associated to the $\mathsf{App}$ constructor. 
% ListC case 
Let CListC be the type 
\begin{align*}
  &\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
  &\mathsf{\to 
  \forall (a \,b : Set) (Q_a : a \to Set)  (Q_b : b \to Set) 
    (e : Equal\, a\, (List\, b)) (ts : List\, (LTerm\, b))} \\ 
  &\mathsf{
    \to Equal^{\wedge}\, a\, (List\,b)\, Q_a\, (List^{\wedge}\, b\, Q_b)\, e 
  \to List^{\wedge}\, (LTerm\,b) (P\, b\, Q_b)\, ts} \\
&\mathsf{
  \to P \, a\, Q_a\, (ListC \,b \,e \, ts) }
\end{align*}
associated to the $\mathsf{ListC}$ constructor. 
Then the type of the deep induction principle for $\mathsf{LTerm}$ is,
\begin{multline*}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)}
\mathsf{\to CVar\,P \to CAbs\,P \to CApp\,P \to CListC\,P} \\
\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(t_a : LTerm\,a) \to LTerm^{\wedge}\,a\,Q_a\,t_a \to P\,a\,Q_a\,t_a}
\end{multline*}

@@

{\color{red} Compare with structural induction? or at least state why structural induction fails.} \\
{\color{red} Can get rid of Maybe using non-empty lists and postulates}

@@

So, using deep induction, we want to prove: 
\[
  \mathsf{getTypeProof : \forall \, (a : Set) \to (t : LTerm\,a) \to getType \,a \,t} \\
\]
which we prove by 
\[
  \mathsf{getTypeProof \,a \,t = 
    DILTerm\, (\lambda \, b\, Q_b\, t\, \to getType \,b\, t) \,
    gtVar\,  gtABs\,  gtApp\,  gtListC\,  a\,\, K1\,\, t\, (LTerm^{\wedge}K1\, a\, t)
    }
\]
where $\mathsf{K1 : a \to Set}$ and $\mathsf{LTerm^{\wedge}K1\, a\, t : LTerm^{\wedge}\, a\, K1\, t}$.
Notice that there is no space in $\mathsf{LTerm^{\wedge}K1}$, because
\[
  \mathsf{LTerm^{\wedge}K1 : \forall (a : Set) (t : LTerm A) \to LTerm^{\wedge}\, a\, K1\,t}
\]
is a function that we will define. In addition to defining $\mathsf{LTerm^{\wedge}K1}$,
we also have to give a proof for each constructor $\mathsf{Var, Abs, App, ListC}$:
\begin{multline*}
  \mathsf{gtVar : \forall (a : Set) (Q_a : a \to Set) (s : String) (T_a : LType\, a) 
      \to LType^{\wedge} \, a\, Q_a\, T_a\, \to Maybe\, (LType\, a)}
\end{multline*}

@@

\begin{itemize}
\item reference (correctly) Haskell Symposium paper
\end{itemize}

>>>>

GADTs are expressive enough to represent, for example, the types and terms of the simply typed lambda calculus. 
We will use a variant of the STLC including a base type of booleans, and type constructors for arrow types and list types. 
We define a GADT $\mathsf{LType}$ to encode the set of permissible types for this calculus:

@@

The $\mathsf{TBool}$ constructor represents the type of booleans in the calculus. It is 
defined in terms of $\mathsf{KBool : \forall\, (a : Set) \to Set}$, which simply returns $\mathsf{Bool}$ on any input:
\[
  \mathsf{KBool \, a = Bool}
\]
The lifting $\mathsf{KBool^{\wedge} : \forall (a : Set) \to (a \to Set) \to (KBool \, a) \to Set}$ lifts every predicate to the constantly true predicate: 
\[
  \mathsf{KBool^{\wedge}\, a\, Q_a\, b = \top}
\]

% The $\mathsf{Bool}$ data type is defined as:
% \begin{equation}\label{eq:bool}
% \begin{array}{l}
% \mathsf{data\, Bool\,: Set\,where}\\
%   \mathsf{\;\;\;\;\;\;\;\;true :\, Bool} \\
%   \mathsf{\;\;\;\;\;\;\;\;false:\, Bool} \\
% \end{array}
% \end{equation}
The type system also includes an arrow type constructor, $\mathsf{TArr}$ , which takes two types $\mathsf{b}$ and $\mathsf{c}$
and produces the arrow type $\mathsf{b \to c}$. Similarly, given a type $\mathsf{b}$, we can construct 
the type $\mathsf{List\,b}$ using the $\mathsf{TList}$ constructor.

The term calculus for this type system includes variables, abstraction, application, 
and also includes an introduction rule that takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.
\begin{equation}\label{eq:eq_lterm}
\begin{array}{l}
\mathsf{data\, LTerm\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Var\,\,\,\,\,\,:\, String \to LType\,a \to LTerm\,a} \\
\mathsf{\;\;\;\;\;\;\;\;Abs\,\,\,\,\,\, :\, \exists (b : Set) (c : Set).\,Equal\,a\,(b \to c) \to String 
      \to LType\,b \to LTerm\,c \to LTerm\,a}\\
  \mathsf{\;\;\;\;\;\;\;\;App\,\,\,\,\, :\, \exists (b : Set).\,LTerm (b \to a) \to LTerm\,b \to LTerm\,a}\\
  \mathsf{\;\;\;\;\;\;\;\;ListC\,\, :\, \exists (b : Set).\,Equal\,a\,(\List\,b) \to List\,(LTerm\,b) \to LTerm\,a}
\end{array}
\end{equation}

The type parameter for $\mathsf{LTerm}$ tracks the types of STLC terms. So the type $\mathsf{LTerm\,A}$ 
contains lambda terms of type $\mathsf{A}$. 
Variables are represented as strings and are tagged with their type. 
The $\mathsf{Var}$ constructor takes a variable name and a type for the variable and produces 
a term of the given type. 
The $\mathsf{Abs}$ constructor takes a variable name, a type $\mathsf{b}$ for the variable,
and a lambda term of type $\mathsf{c}$ for the body of the abstraction, and 
it produces a term of type $\mathsf{b \to c}$. 
The $\mathsf{App}$ constructor applies a lambda term of type $\mathsf{b \to a}$ to a lambda term of type $\mathsf{b}$, producing
a lambda term of type $\mathsf{a}$. 
The $\mathsf{ListC}$ constructor takes a list of lambda terms of type $\mathsf{b}$ and produces a lambda term of type $\mathsf{List\,b}$.

Variables are tagged with their types in the $\mathsf{Var}$ (variable introduction) and 
$\mathsf{Abs}$ (abstraction) constructors. The role of $\mathsf{LType}$ in the $\mathsf{Var}$ and $\mathsf{Abs}$ constructors
is to enforce that variables can only be tagged with a legal type, e.g., $\mathsf{Bool}$, 
$\mathsf{Bool \to Bool}$, $\mathsf{List\, (Bool \to Bool)}$, etc. 
Therefore the presence of $\mathsf{LType}$ in these constructors ensures that all lambda terms produced by 
$\mathsf{Var}$, $\mathsf{Abs}$, $\mathsf{App}$, and $\mathsf{ListC}$ are well-typed.

@@

With these liftings, we can define the deep induction princple for $\mathsf{LTerm}$.
% Var case 
Let CVar be the type 
\begin{multline*}
\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
\mathsf{\to 
  \forall (a : Set) (Q_a : a \to Set) (s : String) (T_a : LType\, a) 
  \to LType^{\wedge} \, a\, Q_a\, T_a
  \to P \, a\, Q_a\, (Var \, s\, T_a)
  }
\end{multline*}
associated to the $\mathsf{Var}$ constructor. 
% Abs case 
Let CAbs be the type 
\begin{align*}
  &\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
  &\mathsf{\to 
  \forall (a \,b\, c: Set) (Q_a : a \to Set)  (Q_b : b \to Set) (Q_c : c \to Set)
  (e : Equal\, a\, (b \to c)) (s : String) } \\ 
  &\mathsf{ \to (T_b : LType\, b) \to (t_c : LTerm\, c)
  \to Equal^{\wedge}\,a\,(b \to c)\, Q_a \, (Arr^{\wedge} \, b\, c\, Q_b \, Q_c) \, e
  } \\
  &\mathsf{
  \to LType^{\wedge}\, b\, Q_b\, T_b
  \to P\, c\, Q_c\, t_c\, 
  \to P \, a\, Q_a\, (Abs \,b \,c \, e \,s \,T_b \, t_c)
  }
\end{align*}
associated to the $\mathsf{Abs}$ constructor. 
% App case 
Let CApp be the type 
\begin{align*}
  &\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
  &\mathsf{\to 
  \forall (a \,b : Set) (Q_a : a \to Set)  (Q_b : b \to Set) 
   (t_{ba} : LTerm\, (b \to a)) (t_b : LTerm\, b)} \\
  &\mathsf{
  \to P\, (b \to a)\, (Arr^{\wedge} \, b\, a\, Q_b \, Q_a) \, t_{ba} \, 
  \to P\, b\, Q_b\, t_b\, 
  \to P \, a\, Q_a\, (App \,b \,t_{ba} \, t_b) }
\end{align*}
associated to the $\mathsf{App}$ constructor. 
% ListC case 
Let CListC be the type 
\begin{align*}
  &\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)} \\
  &\mathsf{\to 
  \forall (a \,b : Set) (Q_a : a \to Set)  (Q_b : b \to Set) 
    (e : Equal\, a\, (List\, b)) (ts : List\, (LTerm\, b))} \\ 
  &\mathsf{
    \to Equal^{\wedge}\, a\, (List\,b)\, Q_a\, (List^{\wedge}\, b\, Q_b)\, e 
  \to List^{\wedge}\, (LTerm\,b) (P\, b\, Q_b)\, ts} \\
&\mathsf{
  \to P \, a\, Q_a\, (ListC \,b \,e \, ts) }
\end{align*}
associated to the $\mathsf{ListC}$ constructor. 
Then the type of the deep induction principle for $\mathsf{LTerm}$ is,
\begin{multline*}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to LTerm\,a \to Set)}
\mathsf{\to CVar\,P \to CAbs\,P \to CApp\,P \to CListC\,P} \\
\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(t_a : LTerm\,a) \to LTerm^{\wedge}\,a\,Q_a\,t_a \to P\,a\,Q_a\,t_a}
\end{multline*}

@@

{\color{red} Can get rid of Maybe using non-empty lists and postulates}

@@

So, using deep induction, we want to prove: 
\[
  \mathsf{getTypeProof : \forall \, (a : Set) \to (t : LTerm\,a) \to getType \,a \,t} \\
\]
which we prove by 
\[
  \mathsf{getTypeProof \,a \,t = 
    DILTerm\, (\lambda \, b\, Q_b\, t\, \to getType \,b\, t) \,
    gtVar\,  gtABs\,  gtApp\,  gtListC\,  a\,\, K1\,\, t\, (LTerm^{\wedge}K1\, a\, t)
    }
\]
where $\mathsf{K1 : a \to Set}$ is the constantly true predicate:
\[
  \mathsf{K1 \, x = \top}
\]
and $\mathsf{LTerm^{\wedge}K1\, a\, t : LTerm^{\wedge}\, a\, K1\, t}$.
Notice that there is no space in $\mathsf{LTerm^{\wedge}K1}$, because
\[
  \mathsf{LTerm^{\wedge}K1 : \forall (a : Set) (t : LTerm A) \to LTerm^{\wedge}\, a\, K1\,t}
\]
is a function that we will define. In addition to defining $\mathsf{LTerm^{\wedge}K1}$,
we also have to give a proof for each constructor $\mathsf{Var, Abs, App, ListC}$:
\begin{multline*}
  \mathsf{gtVar : \forall (a : Set) (Q_a : a \to Set) (s : String) (T_a : LType\, a) 
      \to LType^{\wedge} \, a\, Q_a\, T_a\, \to Maybe\, (LType\, a)}
\end{multline*}

@@

\begin{itemize}
\item reference (correctly) Haskell Symposium paper
\item reference inspiration for STLC GADT : https://www.seas.upenn.edu/~cis194/spring15/lectures/11-stlc.html
